BOLD Parameter Estimation Using Particle Filters

### 2010-08-15 doc/thesis/particlefilters.tex
another pass through single voxel sim

\section{Derivation}
The goal of the particle filter is to evolve an empirical distribution 
$P(x_k | u_{0:k}, Y_{0:k})$,
that asymptotically approaches the true probability distribution $P(X_k | u_{0:k})$.
Note that capital $X$ will be used as the actual realizations of 
the state variable, whereas $x$ will denote estimates of $X$.
Additionally, the notation $a:b$ indicates the set $[a,b]$,
as in $u_{a:b}$, which would indicate all the inputs from time $a$ to time $b$.
Considering the noise present in $X$,
 $P(X_k | u_{0:k})$ is not a single true value but probability distribution.

@@

\section{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$. For practical applications though it is a helpful assumption.
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
The benefit of this choice for importance density is that an approximation for
$P(x_k | x_{k-1})$ is freely available: its simply the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen in \autoref{eq:weightevolve}.

@@

Thus, by making the following simple assumptions, evolving a posterior
density  requires no knowledge of noise distribution.
\begin{enumerate}
\item $f(t, x(t), u(t)) = f(x(t), u(t))$ and $g(t, x(t), u(t)) = g(x(t))$ 
\item The PDF $q(x_i(0))$ (the prior) fully covers $P(x_i(0))$
\item Markov Property: $P(x_k | x_{0:k-1}) = Pr(x_k | x_{k-1})$
\item $q(x_{0:k-1} | y_{0:k}) = q(x_{0:k-1} | y_{0:k-1})$
\end{enumerate}

\subsection{Basic Particle Filter Algorithm}
From the definition of $w_i$, the algorithm to calculate
an approximation of $P(X(t_k) | Y_{0:k})$ or $P(X(t_k + \delta t) | Y_{0:k})$
is simple.

\begin{algorithm}
\caption{Sequential Importance Sampling}
\begin{algorithmic}
\STATE Initialize Particles:
\FOR{$i$ : each of $N_p$ particles }
    \STATE $x^i_0  \sim \alpha(X)$
    \STATE $w^i_0 = \frac{1}{N_p}$
\ENDFOR
\FOR{$k$ : each measurement}
    \FOR{$i$ : each particle }
        \STATE $x^i_k = x^i_{k-1} + \int_{t-1}^t f(x(\tau), u(\tau)) d\tau $
        \STATE $w^i_k = w^i_{k-1}P(y_k | x_k)$
    \ENDFOR
\ENDFOR
\STATE $P(x(t_k+\Delta t)) \approx 
\sum_{i=0}^{N_p} w^i_k \delta\left(x - (x^i_k + \int_{t_k}^{t_k+\Delta t} f(x(\tau), u(\tau)) d\tau) \right)$
\end{algorithmic}
\end{algorithm}

\section{Resampling}
\label{sec:Particle Filter Resampling}
As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), in short order there will be a significant proportion of particles 
with insignificant weights. 
While this does help
describe the tails of the distribution, it means a great deal of computation will be wasted.
Instead, it would be preferable if most of the computation is spent on the most probable regions.
Ideally the computation time spent on tails would be proportional to the actual size of the
tails. In this case particle locations would match the true posterior and all weights would
be equal.  The case where a large number of the weights have become irrelevantly small
is called particle degeneracy. In  \cite{Liu1998b}
an ideal calculation of the effective number of particles is found based on the 
particles' true weight. However, given that only an approximation for the true weight 
exists, they also provide a simple heuristic calculation of $N_{eff}$.
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
Any quick run of a particle filter will reveal that unless the prior is particularly accurate,
$N_{eff}$ drops precipitously.  To alleviate this problem
a common technique known as resampling may be applied. The idea of resampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a better support. Therefore, a new set of particles may be drawn from the empirical
distribution as follows:
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}

>>>>

\section{Sequential Importance Sampling}
The goal of the particle filter is to evolve an empirical distribution 
$P(x_k | u_{0:k}, Y_{0:k})$,
that asymptotically approaches the true probability distribution $P(X_k | u_{0:k})$.
Note that capital $X$ will be used as the actual realizations of 
the state variable, whereas $x$ will denote estimates of $X$.
Additionally, the notation $a:b$ indicates the set $[a,b]$,
as in $u_{a:b}$, which would indicate all the inputs from time $a$ to time $b$.
Considering the noise present in $X$,
 $P(X_k | u_{0:k})$ is not a single true value but probability distribution.

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$. For practical applications though it is a helpful assumption.
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
The benefit of this choice for importance density is that an approximation for
$P(x_k | x_{k-1})$ is freely available: its simply the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen in \autoref{eq:weightevolve}.

@@

Thus, by making the following simple assumptions, evolving a posterior
density  requires no knowledge of noise distribution.
\begin{enumerate}
\item $f(t, x(t), u(t)) = f(x(t), u(t))$ and $g(t, x(t), u(t)) = g(x(t))$ 
\item The prior distribution PDF, $q(x_i(0))$, covers $P(x_i(0))$
\item Markov Property: $P(x_k | x_{0:k-1}) = Pr(x_k | x_{k-1})$
\item $q(x_{0:k-1} | y_{0:k}) = q(x_{0:k-1} | y_{0:k-1})$
\end{enumerate}

\subsection{Basic Algorithm}
From the definition of $w_i$, the algorithm to calculate
an approximation of $P(X(t_k) | Y_{0:k})$ or $P(X(t_k + \delta t) | Y_{0:k})$
is simple. This basic form of the particle filter is given in 
algorithm \autoref{alg:BasicParticleFilter}.

\begin{algorithm}
\caption{Sequential Importance Sampling}
\begin{algorithmic}
\STATE Initialize Particles:
\FOR{$i$ : each of $N_p$ particles }
    \STATE $x^i_0  \sim \alpha(X)$
    \STATE $w^i_0 = \frac{1}{N_p}$
\ENDFOR
\FOR{$k$ : each measurement}
    \FOR{$i$ : each particle }
        \STATE $x^i_k = x^i_{k-1} + \int_{t-1}^t f(x(\tau), u(\tau)) d\tau $
        \STATE $w^i_k = w^i_{k-1}P(y_k | x_k)$
    \ENDFOR
\ENDFOR
\STATE $P(x(t_k+\Delta t)) \approx 
\sum_{i=0}^{N_p} w^i_k \delta\left(x - (x^i_k + \int_{t_k}^{t_k+\Delta t} f(x(\tau), u(\tau)) d\tau) \right)$
\end{algorithmic}
\label{alg:BasicParticleFilter}
\end{algorithm}

\section{Sequential Importance Resampling}
\label{sec:Particle Filter Resampling}
As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), in short order there will be a significant proportion of particles 
with insignificant weights. 
While this does help
describe the tails of the distribution, it means a great deal of computation will be wasted.
Instead, it would be preferable if most of the computation is spent on the most probable regions.
Ideally the computation time spent on tails would be proportional to the actual size of the
tails. In this case particle locations would match the true posterior and all weights would
be equal.  The case where a large number of the weights have become irrelevantly small
is called particle degeneracy. In  \cite{Liu1998b}
an ideal calculation of the effective number of particles is found based on the 
particles' true weight. However, given that only an approximation for the true weight 
exists, they also provide a simple heuristic calculation of $N_{eff}$.
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
Any quick run of a particle filter will reveal that unless the prior is particularly accurate,
$N_{eff}$ drops precipitously.  To alleviate this problem
a common technique known as resampling may be applied. The idea of resampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a better support. Therefore, a new set of particles may be drawn from the empirical
distribution as follows:
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}
